% This is part of the book TeX for the Impatient.
% Copyright (C) 2003 Paul W. Abrahams, Kathryn A. Hargreaves, Karl Berry.
% See file fdl.tex for copying conditions.

\input macros
%\chapter{Using this book}
\chapter{使用本書}

\chapterdef{usebook}

%This book is a do-it-yourself guide and handbook for \TeX. 
%Here in this section we tell you how to use the book to maximum advantage.

本書是一本介紹 \TeX{} 的自學手冊。
在本章中, 我們將指导你如何更好地使用本書。

% We recommend that you first either read or skim in sequence Sections
% \chapternum{usebook} through \chapternum{examples},
% which tell you what you need to know in order to get started using \TeX.
% If you've already had experience using \TeX, it will still be helpful
% to know what kinds of information are in these sections of the book.
% Sections~\chapternum{concepts}--\chapternum{tips}, which
% occupy most of
% the rest of the book, are designed to be accessed randomly.
% Nevertheless, if you're the kind of person who likes to read reference manuals,
% you'll find that it \emph{is} possible to proceed sequentially if you're
% willing to take a lot of detours at first.

我們建議你先閱讀或瀏覧 \chapternum{usebook} 到 \chapternum{examples}
章，
這些地方會告訴你如何上手 \TeX。
如果你已經使用過 \TeX， 這些地方仍會值得一看，
至少你可以知道這些章節介紹了些什麼內容。
%本書的重頭戲是\chapternum{concepts}--\chapternum{tips}章,
%這些章節可以任意選擇地進行閱讀.
其馀部分，主要是\chapternum{concepts}--\chapternum{tips}章，
可以有選擇地翻阅。
當然， 如果你是一個很喜愛閱讀参考手冊的人，
你會發現從頭到尾地把這本書閱讀一遍\emph{是}完全可行的，
當然你會為此在開頭學習時繞些彎路。

% In \chapterref{usingtex}, ``Using \TeX'',
% we explain how to produce a \TeX\ document from a
% \TeX{} input file.
% We also describe the conventions for preparing that input file,
% explain a little about how \TeX{} works, and tell you about additional
% resources that are available.
% Reading this section will help you understand
% the examples in the next section.

在 \chapterref{usingtex}， ``使用 \TeX''，
我們解释如何把一個 \TeX{} 輸入文件轉為一個 \TeX{} 文檔。
%我們還講述了準備這個輸入文件的慣例，
我們還講述了準備輸入文件的一些習慣，
簡要地解釋了 \TeX{} 是如何工作的， 告訴你可以獲取的其它資源信息。
閱讀了這個章節， 會有助于你理解下一個章節的例子。

% \chapterref{examples}, ``Examples'',
% contains a sequence of 
% examples that illustrate the use of \TeX.
% Each example consists of a page of output
% together with the input that we used to create it.
% These examples will orient you and help you 
% locate the more detailed material that you'll need as you go.  
% By seeing which commands are used in the input, you'll know
% where to look for more detailed information on how to achieve
% the effects shown in the output.
% The
% examples can also serve as models for simple documents, although we must
% caution you that 
% because we've tried to pack a variety of \TeX\ commands into a small
% number of pages,  
% the examples are not necessarily illustrations of good or complete 
% document~design.

\chapterref{examples}, ``例子'',
包括了一系列的例子來描述 \TeX{} 的功用.
每個例子包括了一頁的輸出內容, 也包括了輸出該頁內容所輸入的源代碼.
這些例子會指導你且幫助你找到你所需要的更詳細的資料.
通過觀察輸入使用了什麼命令, 
你會知道在哪裡能夠找到輸出效果所使用方法的更詳細的信息.
這些例子也可以作為簡單的文檔的模版, 當然我們必需提醒你, 
因為我們嘗試把很多的 \TeX{} 命令放在這幾頁紙中, 
這些例子不可能描述好的或者完整的文檔設計.

% As you read the explanation of a command, you may encounter
% some unfamiliar technical terms.
% In \chapterref{concepts}, ``Concepts'',
% we define and explain these terms.
% We also discuss other topics that aren't covered
% elsewhere in the book.
% The inside back cover of the book contains a list of all the
% concepts and the pages where they are described.
% We suggest that you make a copy of this list and keep it nearby
% so that you'll be able to identify and look up an unfamiliar
% concept immediately.

當你看這些命令的解釋時, 你會遇到很多不熟悉的技術術語.
在 \chapterref{concepts}, ``概念''中, 我們定義並解釋了這些術語.
我們同時還討論了本書其它地方沒有涉及到的話題.
本書書背內頁列出了所有的概念名稱以及它們所出現的位置.
我們建議你把書背內頁復印下來, 放在旁邊,
這樣你就能在遇到不熟悉的概念時馬上查找到它們.

% \TeX's commands are its primary vocabulary,
% and the largest part of this book is
% devoted to explaining them.  In Sections~\chapternum{paras}
% through~\chapternum{general} we describe the commands.
% You'll find general information about the command descriptions
% on \xrefpg{cmddesc}.
% The command descriptions are arranged
% functionally, rather like a thesaurus, so if you know what you want to
% do but you don't know which command does it for you, you can use the
% table of contents to guide you to the right group of commands.
% Commands that we think are both particularly useful and easy to understand
% are indicated with a pointing~hand~(\hand).

\TeX{} 的命令是它的主要詞彙, 本書最主要的部分就是用來解釋這些命令.
從第~\chapternum{paras}~章到第~\chapternum{general}~章, 我們解釋了這些命令.
你可以在 \xrefpg{cmddesc} 找到命令描述的描述慣例.
這些命令根據功能分門別類, 而不是按字典順序, 
所以如果你知道你想做什麼, 但不知那該用什麼命令,
你可以使用目錄來引導你找到正確的命令組.
我們給我們認為常用且容易理解的命令標上了一個手形符號~(\hand).


\chapterref{capsule}, ``Capsule summary of commands'', is a
specialized index that complements the more complete descriptions
in Sections~\chapternum{paras}--\chapternum{general}.
It lists \TeX's commands 
alphabetically, with a brief explanation of each command
and a reference to the page
where it is described more completely.  The capsule summary
will help you when you just want a quick reminder of what a command
does.

\TeX{} is a complex program that occasionally works its will in
mysterious ways.
In \chapterref{tips}, ``Tips and techniques'',
we provide advice on solving a variety of specific 
problems that you may encounter from time to time.
And if you're stumped by
\TeX's error messages, you'll find succor in \chapterref{errors},
``Making sense of error messages''.

The gray tabs on the side of the book will help you locate parts of the
book quickly.  They divide the book into the following major parts:
\olist
\li general explanations and examples
\li concepts
\li descriptions of commands (five shorter tabs)
\li advice, error messages, and the |eplain.tex| macros
\li capsule summary of commands
\li index
\endolist

In many places we have provided page references to
^{\texbook} (see \xrefpg{resources} for a citation).
These references apply to the seventeenth edition of \texbook.
For other editions, some references may be off by a page or two.


\section Syntactic conventions

In any book about preparing input for a computer,
it's necessary to indicate clearly the literal characters that should be typed
and to distinguish those characters from the explanatory text.
We use the Computer Modern typewriter font for {\tt literal input
like this}, and also for the names of \TeX{} commands.
When there's any possibility of confusion, we enclose \TeX{}
input in single quotation marks, `{\tt like this}'.
However, we occasionally use parentheses when we're indicating single
characters such as (|`|) (you can see why).

For the sake of your eyes we usually just put spaces 
where you should put spaces. In some places where
we need to emphasize the space, however,
we use a `\visiblespace' character
{\catcode `\ =\other\pix^^| |}%
to indicate it.
Naturally enough, this character is called a \emph{visible space}.
\pix^^{spaces//visible}


\section Descriptions of the commands

\xrdef{cmddesc}
Sections~\chapternum{paras}--\chapternum{general} contain
a description of what nearly every \TeX{} command does.  ^^{commands}
Both the primitive commands ^^{primitive//command}
and those of ^{\plainTeX} are covered.
The primitive commands are those built into the \TeX{} computer program, while
the \plainTeX{} commands are defined in a standard file of 
auxiliary definitions (see \xref\plainTeX). 
The only commands we've omitted are those that are used purely locally
in the definition of \plainTeX\ (\knuth{Appendix~B}).
The commands are organized as follows:
\ulist\compact
\li ``Commands for composing paragraphs'', \chapterref{paras},
deal with characters, words, lines, and entire paragraphs.
\li ``Commands for composing pages'', \chapterref{pages},
deal with pages, their components, and the output routine.
\li ``Commands for horizontal and vertical modes'', \chapterref{hvmodes},
have corresponding or identical
forms for both horizontal modes (paragraphs and hboxes) and vertical
modes (pages and vboxes).
These commands provide boxes, spaces, rules, leaders,
and alignments.
\li ``Commands for composing math formulas'', \chapterref{math},
provide capabilities for constructing math formulas.
\li ``Commands for general operations'', \chapterref{general},
provide 
\TeX's programming features and
everything else that doesn't fit into any of the other sections.
\endulist
You should think of these categories as being suggestive rather than
rigorous, because the commands don't really fit neatly into these
(or any other) categories.
 
Within each section, the descriptions of the commands are organized
by function.  When several commands are closely related, they are described as
a group; otherwise, each command has its own explanation.
The description of each command
includes one or more examples and the output
produced by each example when examples are appropriate (for
some commands they aren't).
When you are looking at a subsection containing functionally related
commands, be sure to check the end of a subsection for a ``see also'' 
item that refers you to related commands that are described elsewhere.
 
Some commands are closely related to certain concepts.
For instance, the |\halign| and |\valign| commands are related to
``alignment'', the |\def|
command is related to ``macro'',
and the |\hbox| and |\vbox| commands are related to ``box''.
In these cases we've usually given a bare-bones des\-crip\-tion of the
commands themselves and explained  the underlying ideas
in the concept. 

The examples associated with the commands have been typeset with
^|\parindent|, the paragraph indentation, set to zero so that
paragraphs are normally unindented.
This convention makes the examples easier to read.
In those examples where the paragraph indentation is essential,
we've set it explicitly to a nonzero value.

The pointing hand in front of a command or a group of commands indicates
that we judged this command or group of commands to be particularly useful
and easy to understand.

Many commands expect ^{arguments} of one kind or another
(\xref{arg1}).  The arguments of
a command give \TeX{} additional information that it needs in order to
carry out the command.  Each argument is indicated by an italicized
term in angle brackets that indicates what kind of argument it~is:
 
\display{%
\halign{\<#>\quad&#\hfil\cr
argument&a single token or some text enclosed in braces\cr
charcode&a character code, i.e., an integer between $0$ and $255$\cr
dimen&a dimension, i.e., a length\cr
glue&glue (with optional stretch and shrink)\cr
number&an optionally signed integer (whole number)\cr
register&a register number between $0$ and $255$\cr
}}
^^{\<dimen>}
^^{\<argument>}
^^{\<charcode>}
^^{\<glue>}
^^{\<number>}
^^{\<register>}

% \noindent
% All of these terms are explained in more detail in \chapterref{concepts}.
% In addition, we sometimes use terms such as \<token list> that are either
% self-explanatory or explained in the description of the command.
% Some commands have special formats that require either braces or 
% particular words.
% These are set in the same bold font that we use
% for the command headings.

\noindent
所有這些術語會在 \chapternum{concepts} 中進行詳述.
除了這些以外, 我們會使用一些不言自明的或者在後文進行講述的述語, 比如 \<token list>.
有些命令有自己獨特的格式, 比如需要括孤或者特定的字詞.
它們會使用和命令標題相同的黑體標出.

% Some commands are parameters (\xref{introparms}) or table entries. 
% ^^{parameters//as commands}
% This is indicated in the command's listing.
% You can either use a parameter as an argument or assign a value to it.
% The same holds for table entries.
% We use the term ``parameter'' to refer to entities such as |\pageno|
% that are actually registers but behave just like parameters.
% ^^{registers//parameters as}

有些命令是參項 (\xref{introparms}) 或表項.
^^{parameters//as commands}
這會在命令的列表中指示出來.
你可以作為一個命令的參數使用這個參項, 也可以給它賦值.
表項也是如此.
我們使用``參項''這個術語來指代這些條目, 比如本是寄存器但可當參項使用的 |\pageno|.
^^{registers//parameters as}


\endchapter
\byebye
